Simple repression draft

### 2020-09-24 doc/resubmission/appendix_01_nonbursty.tex
finished transfering main text to SI for appendix 1

\mrm{THIS NEEDS TO BE FIXED}

@@

\subsubsection{The Poisson Promoter Nonequilibrium Model}
For our first kinetic model, we  imitate the states considered in the Two-State
Equilibrium Model and consider the simplest possible picture with only two
states, repressor bound and unbound. This is exactly the model used for the main
results of~\cite{Jones2014}. In this picture, repressor association and
dissociation rates from its operator site, $k_R^+$ and $k_R^-$, respectively,
govern transitions between the two states. When the system is in the unbound
state, transcription initiates at rate $r$, which represents a coarse-graining
of all the downstream processes into a single effective rate. mRNA is degraded
at rate $\gamma$ as already exploited in the previous models.

Let $p_R(m,t)$ denote the joint probability of finding the system in the
repressor bound state $R$ with $m$ mRNA molecules present at time $t$. Similarly
define $p_U(m,t)$ for the repressor unbound state $U$. This model is governed
by coupled master equations giving the time evolution of $p_R(m,t)$ and
$p_U(m,t)$~\cite{Sanchez2008, Sanchez2011, Phillips2019} which we can write as
\begin{align}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma p_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_U(m,t) =&\; 
\overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{rp_U(m-1,t)}^{m-1 \rightarrow m}
- \overbrace{rp_U(m,t)}^{m \rightarrow m + 1}
\\
&+ \overbrace{(m+1)\gamma p_U(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma p_U(m,t)}^{m \rightarrow m - 1},
\label{eq:poisson_promoter_cme}
\end{split}
\end{align}
where each term on the right corresponds to a transition between two states of
the promoter as indicated by the overbrace label. In each equation, the first
two terms describe transitions between promoter states due to repressors
unbinding and binding, respectively. The final two terms describe degradation of
mRNA, decreasing the copy number by one, and the terms with coefficient $r$
describe transcription initiation increasing the mRNA copy number by one.
We direct the reader to Appendix~\ref{sec:cme_from_cartoon} for a careful
treatment showing how the form of this master equation follows from the
corresponding cartoon in Figure~\ref{fig1:means_cartoons}.

We can greatly simplify the notation, which will be especially useful for the
more complicated models yet to come, by re-expressing the master equation in
vector form~\cite{Phillips2012}. The promoter states are collected into a vector
and the rate constants are collected into matrices as
\begin{equation}
\vec{p}(m) = \begin{pmatrix} p_R(m) \\ p_U(m) \end{pmatrix},\
\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix},\
\mathbf{R} = \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix},\
\label{eq:2state_cme_matrices}
\end{equation}
so that the master equation may be condensed as
\begin{equation}
\deriv{t}\vec{p}(m,t) =
\left( \mathbf{K} - \mathbf{R} - \gamma m \mathbf{I} \right) \vec{p}(m,t)
                + \mathbf{R} \vec{p}(m-1,t)
                + \gamma (m+1) \mathbf{I} \vec{p}(m+1,t),
\label{eq:2state_rep_cme}
\end{equation}
where $\mathbf{I}$ is the identity matrix. Taking steady state by setting time
derivatives to zero, the mean mRNA can be found to be
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
\label{eq:mean_m_model1}
\end{equation}
with the algebra details again deferred to Appendix~\ref{sec:non_bursty}. Recall
$k_R^+$ is proportional to the repressor copy number, so in computing
fold-change, absence of repressor corresponds to $k_R^+\rightarrow0$. Therefore
fold-change in this model is simply
\begin{equation}
\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
\end{equation}
again matching the master curve of~\fig{fig1:means_cartoons}(D) with $\rho=1$.

\subsubsection{Nonequilibrium Model Two - RNAP Bound and Unbound States}
Our second kinetic model depicted in Figure~\ref{fig1:means_cartoons}(C) mirrors
the second equilibrium model of Figure~\ref{fig1:means_cartoons}(B) by
fine-graining  the repressor unbound state of nonequilibrium model 1, resolving
it into an empty promoter state and an RNAP-bound state. Note in this picture,
in contrast with model 4 below, transcription initiation is accompanied by a
promoter state change, in keeping with the interpretation as RNAP-bound and
empty states: if an RNAP successfully escapes the promoter and proceeds to
elongation of a transcript, clearly it is no longer bound at the promoter.
Therefore another RNAP must bind before another transcript can be initiated.

The master equation governing this model is analogous
to Eqs.~\ref{eq:2state_cme_matrices}-\ref{eq:2state_rep_cme} for model 1 above. The
main subtlety arises since transcription initiation accompanies a promoter state
change. This can be understood by analogy to $\mathbf{K}$. The off-diagonal and
diagonal elements of $\mathbf{K}$ correspond to transitions arriving at or
departing from, respectively, the promoter state of interest. If transcription
initiation is accompanied by promoter state changes, we must have separate
matrices for arriving and departing transcription events since the arriving and
departing transitions have different initial copy numbers of mRNA, unlike for
$\mathbf{K}$ where they are the same (see Appendix~\ref{sec:non_bursty}). The
master equation for this model is
\begin{equation}
\deriv{t}\vec{p}(m,t) =
\left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m,t)
                + \mathbf{R_A} \vec{p}(m-1,t) +
                \gamma (m+1) \mathbf{I} \vec{p}(m+1,t),
\label{eq:3state_rep_cme}
\end{equation}
with the state vector and promoter transition matrix defined as
\begin{equation}
\vec{p}(m) = \begin{pmatrix} p_R(m) \\ p_E(m) \\ p_P(m) \end{pmatrix},\
\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
                        0 & k_P^+ & -k_P^- 
                \end{pmatrix},
\label{eq:3state_cme_matrices_pt1}
\end{equation}
and the initiation matrices given by
\begin{equation}
\mathbf{R_A} = \begin{pmatrix}
                0 & 0 & 0 \\ 
                0 & 0 & r \\ 
                0 & 0 & 0
                \end{pmatrix},\
\mathbf{R_D} = \begin{pmatrix}
                0 & 0 & 0 \\ 
                0 & 0 & 0 \\ 
                0 & 0 & r
                \end{pmatrix}.
\label{eq:3state_cme_matrices_pt2}
\end{equation}
The elements of $\vec{p}(m)$ encode the probabilities of having $m$ mRNA present
along with the promoter having repressor bound ($R$), being empty ($E$), or
having RNAP bound ($P$), respectively. $\mathbf{R_A}$ describes probability flux
\textit{arriving} at the state $\vec{p}(m)$ from a state with one fewer mRNA,
namely $\vec{p}(m-1)$, and $\mathbf{R_D}$ describes probability flux
\textit{departing} from the state $\vec{p}(m)$ for a state with one more mRNA,
namely $\vec{p}(m+1)$. $\mathbf{K}$ is closely analogous to model 1.

Mean mRNA at steady state is found analogously to model 1, with the result
\begin{equation}
\langle m\rangle = \frac{r}{\gamma}
        \frac{k_R^- k_P^+}
        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)},
\label{eq:model2_meanm}
\end{equation}
and with details again deferred to Appendix~\ref{sec:non_bursty}.
Fold-change is again found from the ratio prescribed by Eq.~\ref{eq:fc_def}, from
which we have
\begin{align}
\text{fold-change}
&=      \frac{k_R^- k_P^+}
        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}
        \frac{k_P^+ + k_P^- + r}{k_P^+}
\\
&=      \left(1 + \frac{k_R^+}{k_R^-}
                \frac{k_P^- + r}{k_P^+ + k_P^- + r}
        \right)^{-1}
\\
&=      \left(1 + \frac{k_R^+}{k_R^-}
        \left(1 + \frac{k_P^+}{k_P^- + r}\right)^{-1}
        \right)^{-1},
\end{align}
which follows the master curve of~\fig{fig1:means_cartoons}(D) with $\rho = 1 +
k_P^+/(k_P^- + r)$ as claimed.

\subsubsection{Nonequilibrium Model Three - Multistep Transcription Initiation
and Escape}
One might reasonably complain that the first two ``nonequilibrium'' models we
have considered are straw men. Their steady states necessarily satisfy detailed
balance which is equivalent to thermodynamic equilibrium. Why is this the case?
At steady state there is by definition no net probability flux in or out of each
promoter state, but since the promoter states form a linear chain, there is only
one way in or out of the repressor bound and RNAP bound states, implying each
edge must actually have a net zero probability flux, which is the definition of
detailed balance (usually phrased as equality of forward and reverse transition
fluxes).

Now we consider model 3 in Figure~\ref{fig1:means_cartoons}(C) which allows the
possibility of true nonequilibrium steady-state fluxes through the promoter
states. We point out that this model was considered previously
in~\cite{Mitarai2015} where a comparison was made with model 1 as used
in~\cite{Jones2014}. The authors of~\cite{Mitarai2015} argued that the
additional complexity is essential to properly account for the noise in the mRNA
distribution. We will weigh in on both models later when we consider observables
beyond fold-change.

The expression for mean mRNA is substantially more complicated now, as worked
out in Appendix~\ref{sec:non_bursty} where we find
\begin{equation}
\langle m\rangle = \frac{r}{\gamma}
        \frac{k_R^- k_P^+ k_O}
        {k_R^- [(k_P^+ (k_O + r) + r(k_P^- + k_O)] + k_R^+ r(k_P^- + k_O)},
\label{eq:model3_mean_m}
\end{equation}
which can be simplified to
\begin{equation}
\langle m\rangle
= \frac{r}{\gamma}
\frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}.
\end{equation}
The strategy is to isolate the terms involving the repressor, so that now the
fold-change is seen to be simply
\begin{align}
\text{fold-change}
&= \frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}
        \frac{1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}}
                {\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
\\
&= \left(
        1 + \frac{k_R^+}{k_R^-}
        \left(1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}\right)^{-1}
\right)^{-1},
\end{align}
surprisingly reducing to the master curve of~\fig{fig1:means_cartoons}(D) once
again, with $\rho = 1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}$.

This example hints that an arbitrarily fine-grained model of downstream
transcription steps may still be collapsed to the form of the master curve for
the means given in Figure~\ref{fig1:means_cartoons}(D), so long as the repressor
binding is exclusive with transcriptionally active states. We offer this as a
conjecture, and we suspect that a careful argument using the King-Altman diagram
method~\cite{King1956, Hill1966} might furnish a ``proof.'' Our focus here is
not on full generality but rather to survey an assortment of plausible models
for simple repression  that have been proposed in the literature.

\subsubsection{Nonequilibrium Model Four - ``Active'' and ``Inactive'' States}
Model 4 in Figure~\ref{fig1:means_cartoons}(C) is at the core of the theory
in~\cite{Razo-Mejia2020}. At a glance the cartoon for this model may appear very
similar to model 2, and mathematically it is, but the interpretation is rather
different. In model 2, we interpreted the third state literally as an RNAP-bound
promoter and modeled initiation of a transcript as triggering a promoter state
change, making the assumption that an RNAP can only
make one transcript at a time. In contrast, in the present model the promoter
state does \textit{not} change when a transcript is initiated. So we no longer
interpret these states as literally RNAP bound and unbound but instead as
coarse-grained ``active'' and ``inactive'' states, the details of which we leave
unspecified for now. We will comment more on this model below when we discuss
Fano factors of models.

Carrying out the algebra, the mean mRNA can be found to be
\begin{equation}
\langle m\rangle = \frac{r}{\gamma}
\frac{k_R^- k^+}
{k_R^- k^+ + k_R^- k^- + k_R^+ k^-},
\end{equation}
and the fold-change readily follows,
\begin{align}
\text{fold-change}
&=      \frac{k_R^- k^+}{k_R^- k^+ + k_R^- k^- + k_R^+ k^-}
        \frac{k_R^- k^+ + k_R^- k^-}{k_R^- k^+}
\\
&=      \left(1 + \frac{k_R^+}{k_R^-}
                \left(1 + \frac{k^+}{k^-}\right)^{-1}
        \right)^{-1},
\end{align}
from which we see $\rho = 1 + k^+/k^-$ as shown in~\fig{fig1:means_cartoons}(C).

\subsubsection{Nonequilibrium Model Five - Bursty Promoter}
The final model we consider shown in Figure~\ref{fig1:means_cartoons}(C) is an
intuitive analog to model 1, with just two states, repressor bound or unbound,
and transition rates between them of $k_R^+$ and $k_R^-$. In model 1, when in
the unbound state, single mRNA transcripts are produced as a Poisson process
with some characteristic rate $r$. The current model by contrast produces, at
some Poisson rate $k_i$, \textit{bursts} of mRNA transcripts.
The burst sizes are assumed to be geometrically distributed with
a mean burst size $b$, which we will motivate in Section~\ref{sec:beyond_means}
when we derive this model as a certain limiting case of model 4.

From this intuitive picture and by analogy to model 1, then, it should be
plausible that the mean mRNA level is
\begin{equation}
\langle m\rangle = \frac{k_i b}{\gamma}
        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
\end{equation}
which will turn out to be correct from a careful calculation. For now, we simply
note that just like model 1, the fold-change becomes
\begin{equation}
\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1}
\end{equation}
with $\rho=1$ also like model 1.
We will also see later how this model emerges as a natural limit of model 4.
\mrm{ENDS HERE}

\mrm{THIS NEEDS TO BE FIXED}
\subsubsection{Noise in the Two-State Promoter, RNAP Bound or Unbound.}

Our second model of constitutive transcription posits an architecture in which
the promoter is either empty or bound by RNAP ~\cite{Phillips2015a,
Phillips2019}. Here,  as shown in model 2 of Figure~\ref{fig2:constit_cartoons}(A),
transcription initiation results in a state transition from the bound to the
unbound state, reflecting the microscopic reality that an RNAP that has begun to
elongate a transcript is no longer available at the start site to begin another.
As shown in Appendix~\ref{sec:non_bursty}, the Fano factor in this model is
given by
\begin{align}
    \nu = 1 -
        \frac{r\rate{k}{P}{+}}
            {\left(\rate{k}{P}{+} + \rate{k}{P}{-} + r\right)
             \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)}.
\label{eq:model2_fano}
\end{align}
The problem with this picture is that the Fano factor is always $<1$. To make
 contact with the experimental reality of $\nu>1$ as shown in
 Figure~\ref{fig2:constit_cartoons}(B), clearly some corrections will be needed.
 While this model adds an appealing element of microscopic reality, we are
 forced to reject it as the additional complexity is unable to capture the
 phenomenology of interest. Obviously the promoter state does in fact proceed
 through cycles of RNAP binding, initiating, and elongating, but it seems that
 the super-Poissonian noise in mRNA copy number we want to model must be
 governed by other features of the system.

\subsubsection{Noise in the Three-State Promoter, Multistep Transcription
Initiation and Escape.}

How might we remedy the deficits of model 2? It is known~\cite{DeHaseth1998}
that once RNAP initially binds the promoter region, a multitude of distinct
steps occur sequentially before RNAP finally escapes into the elongation phase.
Perhaps adding some of this mechanistic detail as shown in model 3 of
Figure~\ref{fig2:constit_cartoons}(A) might rescue the previous model. The next
simplest refinement of that model could consider open complex formation and
promoter escape as separate steps rather than as a single effective step. In
other words, we construct model 3 by adding a single extra state to model 2, and
we will label the two RNAP-bound states as the closed and open complexes,
despite the true biochemical details certainly being more complex. For example,
earlier work extended this model by adding an additional repressor bound state
and did not explicitly consider the limit with no repressor that we analyze
here~\cite{Mitarai2015}. Again, our goal here is not a complete accounting of
all the relevant biochemical detail; this is an exploratory search for the
important features that a model needs to include to square with the known
experimental reality of constitutive expression.

Unfortunately, as hinted at in earlier work~\cite{Mitarai2015}, this model too
has Fano factor $\nu<1$. We again leave the algebraic details for
Appendix~\ref{sec:non_bursty} and merely state the result that
\begin{align}
\nu = 1 - \frac{r k_P^+ k_O}{\mathcal{Z}}
\frac{k_P^+ + k_P^- + k_O + r + \gamma}
    {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + r) + \gamma^2},
\label{eq:model3_fano}
\end{align}
where we defined $\mathcal{Z} = r(k_O + k_P^-) + k_P^+(k_O + r)$ for notational
tidiness. This is necessarily less than 1 for arbitrary rate constants.

In fact, we suspect \textit{any} model in which transcription proceeds through a
multistep cycle must necessarily have $\nu<1$. The intuitive argument compares
the waiting time distribution to traverse the cycle with the waiting time for a
Poisson promoter (model 1) with the same mean time. The latter is simply an
exponential distribution. The former is a convolution of multiple exponentials,
and intuitively the waiting time distribution for a multistep process
should be more peaked with a smaller fractional width than a single
exponential with the same mean.
A less disperse waiting time distribution means transcription initations are
more uniformly distributed in time relative to a Poisson process. Hence the
distribution of mRNA over a population of cells should be less variable compared
to Poisson, giving $\nu<1$.
(In Appendix~\ref{sec:non_bursty} we present a more precise version
of the intuitive arguments in this paragraph.)
Regardless of the merits of this model in describing the noise properties of
constitutive transcription initiation, it ultimately fails the simplest
quantitative feature of the data, namely that the Fano factor $> 1$ and hence
we must discard this mechanistic picture and search elsewhere.
\mrm{ENDS HERE}

@@

@@

Having derived an example chemical master equation we now focus on writing a
general matrix form for the kinetic models 1-4 shown in
Figure~\ref{fig1:means_cartoons}(C) in the main text. In each of these four
models, the natural mathematicization of their cartoons is as a chemical master
equation such as Eq.~\ref{eq:2state_rep_cme} for model 1, which for
completeness we reproduce here as
\begin{align}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_U(m,t) =&\; 
\overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{rp_U(m-1,t)}^{m-1 \rightarrow m}
- \overbrace{rp_U(m,t)}^{m \rightarrow m + 1}
\\
&+ \overbrace{(m+1)\gamma p_U(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_U(m,t)}^{m \rightarrow m - 1}.
\label{eq:poisson_promoter_cme_appdx}
\end{split}
\end{align}
Here $p_R(m,t)$ and $p_U(m,t)$ are the probabilities of finding the system with
$m$ mRNA molecules at time $t$ either in the repressor bound or unbound states,
respectively. $r$ is the probability per unit time that a transcript will be
initiated when the repressor is unbound, and $\gamma$ is the probability per
unit time for a given mRNA to be degraded. $k_R^-$ is the probability per unit
time that a bound repressor will unbind, while $k_R^+$ is the probability per
unit time that an unbound operator will become bound by a repressor. Assuming
mass action kinetics, $k_R^+$ is proportional to repressor copy number $R$.

@@

\subsection{Nonequilibrium Model Four - ``Active'' and ``Inactive'' States}
\subsubsection{Mean mRNA}
The mathematical specification of this model is almost identical to model 2.
The matrix $\mathbf{K}$ is identical, as is $\mathbf{R_D}$. The only difference
is that now $\mathbf{R_A}=\mathbf{R_D}$, i.e., both are diagonal, in contrast to
model 2 where $\mathbf{R_A}$ has an off-diagonal element, as in
Eq.~\ref{eq:model2_matrices_appdx}. Then the analog of
Eq.~\ref{eq:model2_K-R+R_for_m0} for finding $\langle{m}^0\rangle$ is
\begin{equation}
\begin{pmatrix} -k_R^- & k_R^+ & 0 \\
        k_R^- & -k_R^+ -k^+ & k^-\\
        0 & k^+ & -k^-
\end{pmatrix}
\begin{pmatrix}
    p_R \\ p_I \\ p_A
\end{pmatrix}
= 0.
\end{equation}
In fact we need not do this calculation explicitly and can instead recycle the
calculation of mean mRNA $\langle{m}\rangle$ from model 2. The matrices are
identical except for the relabeling $k^- \longleftrightarrow (k_P^- + r)$, and
a careful look through the derivation of $\langle{m}\rangle$ for model 2 shows
that the parameters $k_P^-$ and $r$ only ever appear as the sum $k_P^- + r$. So
taking $\langle{m}\rangle$ from model 2, Eq.~\ref{eq:model2_meanm_appdx}, and
relabeling $(k_P^- + r) \rightarrow k^-$ gives us our answer for model four,
simply
\begin{equation}
\langle{m}\rangle = \frac{r}{\gamma}
        \frac{k^+ k_R^-} {k^+ k_R^- + k^- (k_R^+ + k_R^-)}.
\end{equation}

>>>>

@@

@@

% \subsubsection{Nonequilibrium Model Four - ``Active'' and ``Inactive'' States}
% Model 4 in Figure~\ref{fig1:means_cartoons}(C) is at the core of the theory
% in~\cite{Razo-Mejia2020}. At a glance the cartoon for this model may appear very
% similar to model 2, and mathematically it is, but the interpretation is rather
% different. In model 2, we interpreted the third state literally as an RNAP-bound
% promoter and modeled initiation of a transcript as triggering a promoter state
% change, making the assumption that an RNAP can only
% make one transcript at a time. In contrast, in the present model the promoter
% state does \textit{not} change when a transcript is initiated. So we no longer
% interpret these states as literally RNAP bound and unbound but instead as
% coarse-grained ``active'' and ``inactive'' states, the details of which we leave
% unspecified for now. We will comment more on this model below when we discuss
% Fano factors of models.

% Mathematically this model is very similar to models 1 and 2. Like model 1, the
% matrix $R$ describing transcription initiation is diagonal, namely
% \begin{equation}
% \mathbf{R} = \begin{pmatrix}
%                 0 & 0 & 0 \\ 
%                 0 & 0 & 0 \\ 
%                 0 & 0 & r
%         \end{pmatrix}.
% \end{equation}
% The master equation takes verbatim the same form as it did for model 1,
% Eq.~\ref{eq:2state_rep_cme}. Meanwhile the promoter transition
% matrix $\mathbf{K}$ is the same as Eq.~\ref{eq:3state_cme_matrices_pt1}
% from model 2, although we relabel the
% rate constants from $k_P^\pm$ to $k^\pm$ to reiterate that these are not simply
% RNAP binding and unbinding rates.

% Carrying out the algebra, the mean mRNA can be found to be
% \begin{equation}
% \langle m\rangle = \frac{r}{\gamma}
% \frac{k_R^- k^+}
% {k_R^- k^+ + k_R^- k^- + k_R^+ k^-},
% \end{equation}
% and the fold-change readily follows,
% \begin{align}
% \text{fold-change}
% &=      \frac{k_R^- k^+}{k_R^- k^+ + k_R^- k^- + k_R^+ k^-}
%         \frac{k_R^- k^+ + k_R^- k^-}{k_R^- k^+}
% \\
% &=      \left(1 + \frac{k_R^+}{k_R^-}
%                 \left(1 + \frac{k^+}{k^-}\right)^{-1}
%         \right)^{-1},
% \end{align}
% from which we see $\rho = 1 + k^+/k^-$ as shown in~\fig{fig1:means_cartoons}(C).

% \subsubsection{Nonequilibrium Model Five - Bursty Promoter}
% The final model we consider shown in Figure~\ref{fig1:means_cartoons}(C) is an
% intuitive analog to model 1, with just two states, repressor bound or unbound,
% and transition rates between them of $k_R^+$ and $k_R^-$. In model 1, when in
% the unbound state, single mRNA transcripts are produced as a Poisson process
% with some characteristic rate $r$. The current model by contrast produces, at
% some Poisson rate $k_i$, \textit{bursts} of mRNA transcripts.
% The burst sizes are assumed to be geometrically distributed with
% a mean burst size $b$, which we will motivate in Section~\ref{sec:beyond_means}
% when we derive this model as a certain limiting case of model 4.

% From this intuitive picture and by analogy to model 1, then, it should be
% plausible that the mean mRNA level is
% \begin{equation}
% \langle m\rangle = \frac{k_i b}{\gamma}
%         \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
% \end{equation}
% which will turn out to be correct from a careful calculation. For now, we simply
% note that just like model 1, the fold-change becomes
% \begin{equation}
% \text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1}
% \end{equation}
% with $\rho=1$ also like model 1.
% We will also see later how this model emerges as a natural limit of model 4.
% \mrm{ENDS HERE}

% \mrm{THIS NEEDS TO BE FIXED}
% \subsubsection{Noise in the Two-State Promoter, RNAP Bound or Unbound.}

% Our second model of constitutive transcription posits an architecture in which
% the promoter is either empty or bound by RNAP ~\cite{Phillips2015a,
% Phillips2019}. Here,  as shown in model 2 of Figure~\ref{fig2:constit_cartoons}(A),
% transcription initiation results in a state transition from the bound to the
% unbound state, reflecting the microscopic reality that an RNAP that has begun to
% elongate a transcript is no longer available at the start site to begin another.
% As shown in Appendix~\ref{sec:non_bursty}, the Fano factor in this model is
% given by
% \begin{align}
%     \nu = 1 -
%         \frac{r\rate{k}{P}{+}}
%             {\left(\rate{k}{P}{+} + \rate{k}{P}{-} + r\right)
%              \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)}.
% \label{eq:model2_fano}
% \end{align}
% The problem with this picture is that the Fano factor is always $<1$. To make
%  contact with the experimental reality of $\nu>1$ as shown in
%  Figure~\ref{fig2:constit_cartoons}(B), clearly some corrections will be needed.
%  While this model adds an appealing element of microscopic reality, we are
%  forced to reject it as the additional complexity is unable to capture the
%  phenomenology of interest. Obviously the promoter state does in fact proceed
%  through cycles of RNAP binding, initiating, and elongating, but it seems that
%  the super-Poissonian noise in mRNA copy number we want to model must be
%  governed by other features of the system.

% \subsubsection{Noise in the Three-State Promoter, Multistep Transcription
% Initiation and Escape.}

% How might we remedy the deficits of model 2? It is known~\cite{DeHaseth1998}
% that once RNAP initially binds the promoter region, a multitude of distinct
% steps occur sequentially before RNAP finally escapes into the elongation phase.
% Perhaps adding some of this mechanistic detail as shown in model 3 of
% Figure~\ref{fig2:constit_cartoons}(A) might rescue the previous model. The next
% simplest refinement of that model could consider open complex formation and
% promoter escape as separate steps rather than as a single effective step. In
% other words, we construct model 3 by adding a single extra state to model 2, and
% we will label the two RNAP-bound states as the closed and open complexes,
% despite the true biochemical details certainly being more complex. For example,
% earlier work extended this model by adding an additional repressor bound state
% and did not explicitly consider the limit with no repressor that we analyze
% here~\cite{Mitarai2015}. Again, our goal here is not a complete accounting of
% all the relevant biochemical detail; this is an exploratory search for the
% important features that a model needs to include to square with the known
% experimental reality of constitutive expression.

% Unfortunately, as hinted at in earlier work~\cite{Mitarai2015}, this model too
% has Fano factor $\nu<1$. We again leave the algebraic details for
% Appendix~\ref{sec:non_bursty} and merely state the result that
% \begin{align}
% \nu = 1 - \frac{r k_P^+ k_O}{\mathcal{Z}}
% \frac{k_P^+ + k_P^- + k_O + r + \gamma}
%     {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + r) + \gamma^2},
% \label{eq:model3_fano}
% \end{align}
% where we defined $\mathcal{Z} = r(k_O + k_P^-) + k_P^+(k_O + r)$ for notational
% tidiness. This is necessarily less than 1 for arbitrary rate constants.

% In fact, we suspect \textit{any} model in which transcription proceeds through a
% multistep cycle must necessarily have $\nu<1$. The intuitive argument compares
% the waiting time distribution to traverse the cycle with the waiting time for a
% Poisson promoter (model 1) with the same mean time. The latter is simply an
% exponential distribution. The former is a convolution of multiple exponentials,
% and intuitively the waiting time distribution for a multistep process
% should be more peaked with a smaller fractional width than a single
% exponential with the same mean.
% A less disperse waiting time distribution means transcription initations are
% more uniformly distributed in time relative to a Poisson process. Hence the
% distribution of mRNA over a population of cells should be less variable compared
% to Poisson, giving $\nu<1$.
% (In Appendix~\ref{sec:non_bursty} we present a more precise version
% of the intuitive arguments in this paragraph.)
% Regardless of the merits of this model in describing the noise properties of
% constitutive transcription initiation, it ultimately fails the simplest
% quantitative feature of the data, namely that the Fano factor $> 1$ and hence
% we must discard this mechanistic picture and search elsewhere.
% \mrm{ENDS HERE}

@@

Having derived an example chemical master equation we now focus on writing a
general matrix form for the kinetic models 1-4 shown in
Figure~\ref{fig1:means_cartoons}(C) in the main text. In each of these four
models, the natural mathematicization of their cartoons is as a chemical master
equation such as the coupled chemical master equations we derived for model 1,
which for completeness we reproduce here as
\begin{align}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_U(m,t) =&\; 
\overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{rp_U(m-1,t)}^{m-1 \rightarrow m}
- \overbrace{rp_U(m,t)}^{m \rightarrow m + 1}
\\
&+ \overbrace{(m+1)\gamma p_U(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_U(m,t)}^{m \rightarrow m - 1}.
\label{eq:poisson_promoter_cme_appdx}
\end{split}
\end{align}
Here $p_R(m,t)$ and $p_U(m,t)$ are the probabilities of finding the system with
$m$ mRNA molecules at time $t$ either in the repressor bound or unbound states,
respectively. $r$ is the probability per unit time that a transcript will be
initiated when the repressor is unbound, and $\gamma$ is the probability per
unit time for a given mRNA to be degraded. $k_R^-$ is the probability per unit
time that a bound repressor will unbind, while $k_R^+$ is the probability per
unit time that an unbound operator will become bound by a repressor. Assuming
mass action kinetics, $k_R^+$ is proportional to repressor copy number $R$.

@@

\subsection{Kinetic Model Four - ``Active'' and ``Inactive'' States}
Model 4 in Figure~\ref{fig1:means_cartoons}(C) is at the core of the theory
in~\cite{Razo-Mejia2020}. At a glance the cartoon for this model may appear very
similar to model 2, and mathematically it is, but the interpretation is rather
different. In model 2, we interpreted the third state literally as an RNAP-bound
promoter and modeled initiation of a transcript as triggering a promoter state
change, making the assumption that an RNAP can only
make one transcript at a time. In contrast, in the present model the promoter
state does \textit{not} change when a transcript is initiated. So we no longer
interpret these states as literally RNAP bound and unbound but instead as
coarse-grained ``active'' and ``inactive'' states, the details of which we leave
unspecified for now. We will comment more on this model below when we discuss
Fano factors of models.

\subsubsection{Mean mRNA}
The mathematical specification of this model is almost identical to model 2.
The matrix $\mathbf{K}$ is identical, as is $\mathbf{R_D}$. The only difference
is that now $\mathbf{R_A}=\mathbf{R_D}$, i.e., both are diagonal, in contrast to
model 2 where $\mathbf{R_A}$ has an off-diagonal element, as in
Eq.~\ref{eq:model2_matrices_appdx}. Then the analog of
Eq.~\ref{eq:model2_K-R+R_for_m0} for finding $\langle{m}^0\rangle$ is
\begin{equation}
\begin{pmatrix} -k_R^- & k_R^+ & 0 \\
        k_R^- & -k_R^+ -k^+ & k^-\\
        0 & k^+ & -k^-
\end{pmatrix}
\begin{pmatrix}
    p_R \\ p_I \\ p_A
\end{pmatrix}
= 0.
\end{equation}
In fact we need not do this calculation explicitly and can instead recycle the
calculation of mean mRNA $\langle{m}\rangle$ from model 2. The matrices are
identical except for the relabeling $k^- \longleftrightarrow (k_P^- + r)$, and
a careful look through the derivation of $\langle{m}\rangle$ for model 2 shows
that the parameters $k_P^-$ and $r$ only ever appear as the sum $k_P^- + r$. So
taking $\langle{m}\rangle$ from model 2, Eq.~\ref{eq:model2_meanm_appdx}, and
relabeling $(k_P^- + r) \rightarrow k^-$ gives us our answer for model four,
simply
\begin{equation}
\langle{m}\rangle = \frac{r}{\gamma}
        \frac{k^+ k_R^-} {k^+ k_R^- + k^- (k_R^+ + k_R^-)}.
\end{equation}

\subsubsection{Fold-Change}
